\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-07 k-planar.tex

\note{DW}{I suggest we replace $2g+8$ by $2g+6$ here and replace 4 by 3, and cite \citep{DHHW,UWY}. This will improve the bounds and (more importantly) simplify the presentation of various results (such as \cref{StringPartition,PowerGenus}) which currently distinguish the $g=0$ case. }
\note{PM}{I agree with this suggestion.  I've only added the other two references here, and haven't changed anything elsewhere yet.}

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
		\begin{compactenum}
			\item No edge of $F$ is crossed by any edge of $G$; and
			\item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
			\begin{compactenum}
				\item $F[P_i]$ is a path; and
				\item $|V(P_i)\cap L_j| \le 15$ for every integer $j\ge 0$.
			\end{compactenum}
		\end{compactenum}
	\note{DW}{Should the vertices already grabbed by $P_1,P_2,P_3$ be mentioned in the setup (since the new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$)?. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 15$;
		\item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$
		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}; and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

\begin{proof}
	This proof is very similar to the proof of Lemma~14 by \citet{DJMMUW20}. Rather than duplicate every detail of that proof here, we focus on the differences and refer the reader to the original proof for the remaining details.

	The proof is by induction on the number of vertices of $N$.
	First note that $N'$ is a near-triangulation.  If $k=3$, set $R_i := P_i$ for each $i\in\{1,2,3\}$.  Otherwise, as in \citep{DJMMUW20}, split $P_1,\ldots,P_k$ to partition $V(F)$ into three sets $R_1$, $R_2$, and $R_3$ such that each $F[R_i]$ is a non-empty path and each $R_i$ contains vertices from exactly one of $P_1,\ldots,P_k$.

	Next, as in \citep{DJMMUW20}, use Sperner's Lemma to find an inner face $\tau=v_1v_2v_3$ of $N'$ such that, $T$ contains disjoint vertical paths $Q_1,Q_2,Q_3$ such that each $Q_i$ begins at $v_i$, ends at some vertex in $R_i$, and whose internal vertices (if any) are contained in $N'-V(F)$.

	Let $\overline{Y}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $\overline{Y}^+$ denote the subgraph of $N$ consisting of the vertices and edges of $\overline{Y}$ plus the vertices and edges of every kite formed by a crossing between an edge of $G$ and an edge of $\overline{Y}$.

>>>>

\note{DW}{I suggest we replace $2g+8$ by $2g+6$ here and replace 4 by 3, and cite \citep{DHHW,UWY}. This will improve the bounds and (more importantly) simplify the presentation of various results (such as \cref{StringPartition,PowerGenus}) which currently distinguish the $g=0$ case. }

\note{PM}{I agree with this suggestion.  I've only added the other two references here, and haven't changed anything elsewhere yet.}

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}. 
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
		\begin{compactenum}
			\item No edge of $F$ is crossed by any edge of $G$; and
			\item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
			\begin{compactenum}
				\item $F[P_i]$ is a path; and
				\item $|V(P_i)\cap L_j| \le 15$ for every integer $j\ge 0$.
			\end{compactenum}
		\end{compactenum}
	\note{DW}{Should the vertices already grabbed by $P_1,P_2,P_3$ be mentioned in the setup? The new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.}
  \note{PM}{No, at this point we're only promising an $H$-partition of $N$, the part of $G$ contained in the uncrossed cycle $F$.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 15$;
		\item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$
		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}\note{PM}{No, for the same reason as above.}; and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

\begin{proof}
	This proof is very similar to the proof of Lemma~14 by \citet{DJMMUW20}, but is sufficiently complicated by the $1$-planar setting to warrant a full proof.
   % 
   % 
   % 
   % 
   % Rather than duplicate every detail of that proof here, we focus on the differences and refer the reader to the original proof for the remaining details.
  In particular, the reader should keep in mind that $G$ and $G'$ are multigraphs and that the cycle $F$ may consist of two vertices and two (parallel) edges.

  The proof is by induction on the number of inner vertices of $N$, i.e., on $|V(N)\setminus V(F)|$.  If $N$ has no inner vertices, then we can simply take $\mathcal{P}:=\{P_1,\ldots,P_k\}$ and verify the preconditions of the lemma (the setup) ensure that $\mathcal{P}$ satisfies the requirements of the lemma.  Therefore, we may assume that $N$ has at least one inner vertex.

	First note that every inner face of $N'$ is bounded by three edges of $G'$. 
  Assign each vertex $v$ of $N'$ a colour $\alpha(v)\in\{1,2,3\}$ as follows: 
  Let $w$ be the first vertex of $F$ encountered on the path in $T$ from $v$ to the root of $T$.  (This implies that $w=v$ for $v\in V(F)$.)  Since $w\in V(F)$, $w\in P_i$ for some $i\in\{1,2,3\}$ and we define $\alpha(v):=i$.

  If $k<3$ then let $\tau:=v_1v_2v_3$ be any inner face of $N'$.  Otherwise, $k=3$, in which case we apply Sperner's Lemma implies to obtain an inner face $\tau:=v_1v_2v_3$ such that $\alpha(v_i)=i$ for each $i\in\{1,2,3\}$.  
  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to $V(F)$.
   % 
   % 
   % If $k=3$, set $R_i := P_i$ for each $i\in\{1,2,3\}$.  Otherwise, as in \citep{DJMMUW20}, split $P_1,\ldots,P_k$ to partition $V(F)$ into three sets $R_1$, $R_2$, and $R_3$ such that each $F[R_i]$ is a non-empty path and each $R_i$ contains vertices from exactly one of $P_1,\ldots,P_k$.
  % 
	% Next, as in \citep{DJMMUW20}, use Sperner's Lemma to find an inner face $\tau=v_1v_2v_3$ of $N'$ such that, $T$ contains disjoint vertical paths $Q_1,Q_2,Q_3$ such that each $Q_i$ begins at $v_i$, ends at some vertex in $R_i$, and whose internal vertices (if any) are contained in $N'-V(F)$.
	Let $\overline{Y}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $\overline{Y}^+$ denote the subgraph of $N$ consisting of the vertices and edges of $\overline{Y}$ plus the vertices and edges of every kite formed by a crossing between an edge of $G$ and an edge of $\overline{Y}$.
